Parsing Postfix log files

### 2007-05-27 doc/logparser.tex
Started my bibliography - unnecessarily hard in latex, I think.  It's nicely set up now though.
Worked through sections one and two adding bibliography references and doing some TODO items.

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or to check whether
their new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or complaining
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing lines independently.
One of the barriers to this type of processing is the unstructured nature
of Postfix logs, where each logging line was added on an ad hoc basis as a
requirement was discovered or new functionality was added.  Further
complication arises as a result of the facility to have rejection messages
defined by the administrator; every Real-time Black List returns a
different explanatory message; policy servers may have different messages
depending on the characteristics of the connection; there are many ways in
which the log lines may differ.  This paper documents the difficult process
of parsing Postfix logs, presenting a program which parses logs and places
the resulting data into a database.  The gathered data can then be used to
optimise current anti-spam measures, provide a baseline to test new
anti-spam measures against, or to produce statistics showing how effective
those measures are.  There are numerous other uses for such data: improving
server performance by identifying troublesome destinations and
reconfiguring appropriately or dedicating transports to those destinations;
identifying high volume uses (e.g.\ customer newsletters) and restricting
those uses to off-peak times; as a base for billing customers on a shared
server; etc.

Section~\ref{background} provides background information about: the
motivation for this project; the Postfix Mail Transport Agent; a short
overview of the parser; the assumptions inherent in the design of the
algorithm and program; using a database schema as an API\@; and a
comparison against other Postfix log parsers.

@@

Section~\ref{conclusion} describes the results of the project to date.  XXX
IMPROVE THIS\@.

@@

If the database supports sub-selects percentages can be
obtained:\footnote{This works for SQLite3, where || is the concatenation
operator.  If the database containing the data does not support
this syntax, simply remove ``|| '\%\''' from the query; the results will be
the same, just slightly less visually pleasing.}

XXX STOP THIS FROM BEING SPLIT ACROSS PAGES IF POSSIBLE

@@

Another example is determining which restrictions are not effective:

@@

    \item Libraries exist for the majority of programming languages
        allowing access to databases, whereas if this program was written
        as a library a new interface layer would need to be written for
        every programming language used with the API\@.

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive anti-spam restrictions,
allowing a mail administrator to deploy those restrictions which they judge
suitable for their needs, rather than a fixed set chosen by Postfix's
author.  These restrictions can be selectively applied, combined and
bypassed on a per-client, per-recipient or per-sender basis, allowing
varying levels of defense and or permissiveness.  Postfix leverages simple
lookup tables to support arbitrarily complicated user-defined sequences of
restrictions and exceptions, with policy servers as the ultimate in
flexibility.  Administrators can also supply their own rejection messages
to make it clear to senders why exactly their mail was rejected.
Unfortunately this flexibility has a cost: complexity in the logs
generated.  While it is easy to use \texttt{grep(1)} to determine the fate
of an individual email, following the journey an email takes through
Postfix can be quite difficult.  The logs tend to follow a 90\%-10\%
pattern: 90\% of the time the journey is simple, but the other 10\% of the
time requires 90\% of the code.\footnote{These numbers don't have a solid
scientific basis, they're based on gut feeling from writing and debugging
the software.}

A policy server is an external program that accepts state information from
Postfix for each SMTP command not rejected by an earlier restriction; the
policy server can utilise that state information to implement whatever
logic is required.  E.g.\ some users can be restricted to sending mail on
the third Tuesday after pay day only --- this example may actually be
useful in a payroll system to prevent spam or (worse) phishing mails with
faked sender addresses.  More commonly encountered scenarios are:

@@

    \item Checking Sender Policy Framework records.  SPF records specify
        which mail servers are allowed to send mail claiming to be from a
        particular domain.  The intention is to reduce faked sender
        addresses, which will in turn reduce spam, backscatter and
        joe-jobs; however there has been a lot of resistance to the
        proposal because it breaks or vastly complicates some features of
        SMTP.  For more details see \url{http://www.openspf.org/} or
        \url{http://en.wikipedia.org/wiki/Sender_Policy_Framework} XXX ADD
        REFERENCES FOR SPAM BACKSCATTER AND JOE-JOBS TO THE APPENDIX.

    \item Greylisting, a technique that temporarily rejects mail where the
        triple of (sender, recipient, remote IP address) is unknown; the
        assumption is that it's uneconomical for a spammer to retry, but
        that a legitimate mail server must retry.  Sadly spammers are using
        increasingly complex and well written programs to distribute spam,
        frequently using an ISP provided SMTP server from a compromised
        machine on its network, so greylisting is slowly becoming less
        useful, but it does block a large percentage of spam mail at the
        moment.  See \url{http://www.greylisting.org/} for full details.

    \item Using a scoring system such as Policyd-weight where 
        tests accumulate points against the sending system --- if the
        eventual score is too high the mail is rejected.
        \url{http://www.policyd-weight.org/}

    \item Policyd performs the following: \textit{Greylisting,
        Sender-(envelope, SASL or host / ip)-based throttling (on messages
        and/or volume per defined time unit), Recipient rate limiting,
        Spamtrap monitoring / blacklisting, HELO auto blacklisting and HELO
        randomization preventation.\/}  \url{http://policyd.sourceforge.net}

@@

Example attributes taken from \newline
\url{http://www.postfix.org/SMTPD_POLICY_README.html}:

@@

Before getting into detail about the parser a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules can be thought of as the Lex part of
a Lex and YACC style parser; rules identify the line and return some data,
which the algorithm (the YACC part) receives and performs the requested
action.  Rules are solely concerned with identifying a line and extracting
data from it, whereas the algorithm's task is to follow the journey each
mail takes through Postfix, piecing the data together into a coherent
whole, saving it in a useful and consistent form, and performing
housekeeping duties.

XXX REFER TO THE GRAMMAR COMPARISON SECTION ONCE ITS WRITTEN.

@@

XXX DO I NEED TO SAY MORE??

@@

XXX COMPARE AND CONTRAST WITH CONTEXT FREE GRAMMARS, OTHER GRAMMARS.

@@

            \item [save\_by\_pid] Use the pid from the log line to lookup a
                connection, then save the data from the log line in the
                connection data structure thus found.

@@

XXX CARL SUGGESTS EXPANDING ON THIS SOME MORE, BUT I DON'T KNOW HOW\@.
PERHAPS WHEN THE ALGORITHM IS PROPERLY EXPLAINED IT'LL BE EASIER\@?  MAKE
REFERENCES TO THE ALGORITHM, EXPLAIN MORE CLEARLY HOW THE FIELDS ARE USED,
HOW THE RULE MATCHES, AND SO ON\@.

@@

>>>>

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or check whether their
new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or complaining
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing lines independently.
One of the barriers to this type of processing is the unstructured nature
of Postfix logs, where each logging line was added on an ad hoc basis as a
requirement was discovered or new functionality was added.  Further
complication because the list of rejection messages is not fixed: new
messages can be added by the administrator; every Real-time Black List
returns a different explanatory message; policy servers may have different
messages depending on the characteristics of the connection; there are many
ways in which the log lines may differ.  This paper documents the difficult
process of parsing Postfix logs, presenting a program which parses logs and
places the resulting data into a database.  The gathered data can then be
used to optimise current anti-spam measures, provide a baseline to test new
anti-spam measures against, or to produce statistics showing how effective
those measures are.  There are numerous other uses for such data: improving
server performance by identifying troublesome destinations and
reconfiguring appropriately or dedicating transports to those destinations;
identifying high volume uses (e.g.\ customer newsletters) and restricting
those uses to off-peak times; as a base for billing customers on a shared
server; etc.

Section~\ref{background} provides background information about the
following: the motivation for this project; the Postfix Mail Transport
Agent; a short overview of the parser; the assumptions inherent in the
design of the algorithm and program; using a database schema as an API\@;
and a comparison against other Postfix log parsers.

@@

Section~\ref{conclusion} describes the results of the project to date and
contains the paper's conclusion.

@@

If the database supports sub-selects percentages can be
obtained:

@@

SQL note: $||$ is the concatenation operator in SQLite3; if the database
containing the extracted data does not support this syntax, then simply
remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the results will be
the same, just slightly less visually pleasing.

Another example is determining which restrictions are not effective: this
example shows which rules had less than 100 hits on the last log file
parsed.

@@

    \item Libraries exist for the majority of programming languages
        allowing access to databases, whereas if this parser was written
        as a library a new interface layer would need to be written for
        every programming language using API\@.

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive optional anti-spam
restrictions, allowing an administrator to deploy those restrictions which
they judge suitable for their site's needs, rather than a fixed set chosen
by Postfix's author.  These restrictions can be selectively applied,
combined and bypassed on a per-client, per-recipient or per-sender basis,
allowing varying levels of defense and/or permissiveness.  Postfix
leverages simple lookup tables to support arbitrarily complicated
user-defined sequences of restrictions and exceptions, with policy
servers\footnote{Policy servers will be explained shortly.} as the ultimate
in flexibility.  Administrators can also supply their own rejection
messages to make it clear to senders why exactly their mail was rejected.
Unfortunately this flexibility has a cost: complexity in the logs
generated.  While it is easy to use \texttt{grep(1)} to determine the fate
of an individual email, following the journey an email takes through
Postfix can be quite difficult.  The logs tend to follow a 90\%-10\%
pattern: 90\% of the time the journey is simple, but the other 10\% of the
time requires 90\% of the code.\footnote{These numbers don't have a solid
scientific basis, they're based on gut feeling from writing and debugging
the software.}

\subsubsection{Policy servers}

A policy server \cite{policy-servers} is an external program that accepts
state information from Postfix for each SMTP command not rejected by an
earlier restriction; the policy server can utilise that state information
to implement whatever logic is required.  E.g.\ some users can be
restricted to sending mail on the third Tuesday after pay day only --- this
example may actually be useful in a payroll system to prevent spam or
(worse) phishing mails with faked sender addresses.  More commonly
encountered scenarios are:

@@

    \item Checking Sender Policy Framework records \cite{openspf},
        \cite{wikipedia-spf}.  SPF records specify which mail servers are
        allowed to send mail claiming to be from a particular domain.  The
        intention is to reduce spam from faked sender addresses,
        backscatter \cite{postfix-backscatter} and joe-jobs
        \cite{wikipedia-joe-job}; however there
        has been a lot of resistance to the proposal because it breaks or
        vastly complicates some features of SMTP.

    \item Greylisting \cite{greylisting}, a technique that temporarily
        rejects mail where the triple of (sender, recipient, remote IP
        address) is unknown; the assumption is that it's uneconomical for a
        spammer to retry, but that a legitimate mail server must retry.
        Sadly spammers are using increasingly complex and well written
        programs to distribute spam, frequently using an ISP provided SMTP
        server from a compromised machine on its network.  Greylisting is
        slowly becoming less useful, but it does block a large percentage
        of spam mail at the moment.

    \item Using a scoring system such as Policyd-weight
        \cite{policyd-weight} where tests accumulate points against the
        sending system --- if the eventual score is too high the mail is
        rejected.

    \item Rate limiting or throttling on a per-sender, per-client or
        per-recipient basis as performed by Policyd \cite{policyd}.

@@

Example attributes taken from \cite{policy-servers}:

@@

Before getting into detail about the parser a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules can be thought of as the Lex part of
a Lex and YACC style parser; rules identify the line and return some data,
which the algorithm (the YACC part) receives and performs the requested
action.  Rules are solely concerned with identifying a line and extracting
data from it, whereas the algorithm's task is to follow the journey each
mail takes through Postfix, piecing the data together into a coherent
whole, saving it in a useful and consistent form, and performing
housekeeping duties.  Readers familiar with context-free grammars may find
Section \ref{comparison against context-free grammars} helpful.

@@

@@

\subsection{Comparison against context-free grammars}

\label{comparison against context-free grammars}

@@

@@

XXX CARL SUGGESTS EXPANDING ON THIS SOME MORE, BUT I DON'T KNOW HOW\@.
PERHAPS WHEN THE ALGORITHM IS PROPERLY EXPLAINED IT'LL BE EASIER\@?  MAKE
REFERENCES TO THE ALGORITHM, EXPLAIN MORE CLEARLY HOW THE FIELDS ARE USED,
HOW THE RULE MATCHES, AND SO ON\@.  SAVE\_BY\_PID IS GONE NOW, USE A
DIFFERENT EXAMPLE.

@@

% Bibliography.
\bibliographystyle{logparser-bibliography-style}
\bibliography{logparser-bibliography}
